---
title: "hw02-Alina-Skripets"
author: "Alina Skripets"
date: "October 1, 2017"
output: github_document
---

About the Research Question
In the previous assignment you started to tackle the initial research question: “the more
points a player scores, the higher his salary?”
Admittedly, this question is kind of open ended. From a narrow point of view, and based
on the correlation between points and salary, we could answer the research question like so:
“Yes, in general, the more points scored by a player, the higher his salary”. We can even take
one more step and use the results of the simple linear regression to state that “on average,
for every extra point that a player scores, his salary will tend to increase by 8556.68 dollars”.
However, the salary of a player will hardly depend on only the number of scored points.
Other factors like years of experience, offensive performance, defensive performance, and
physical attributes, for instance, may contribute as well to his salary.
In this assignment, you are going to expand the scope of the analysis by taking into account
the overall performance of a player. More precisely, the idea is to look at common individual
statistics (e.g. points, rebounds, assists, steals, etc.) to try to capture ther performance of a
player, and see how this may be related with his salary.

# Data

The data set for this assignment is in the file nba2017-player-statistics.csv, inside the
data/ folder of the course github repo stat133-fall-2017.

# Download the data
You will need to download a copy of the data file to your local repository. One way to do
this from R is with the download.file() function. By the way, you should NOT include a
code chunk with the previous command in your Rmd file.

```{r}
# download csv file to your working directory
# (do NOT include this code in your Rmd)
rm(list = ls())
github <- "https://github.com/ucb-stat133/stat133-fall-2017/raw/master/"
file <- "data/nba2017-player-statistics.csv"
csv <- paste0(github, file)
#download.file(url = csv, destfile = 'nba2017-player-statistics.csv')
```

```{r}

#install.packages("readr")
library(readr)
library(ggplot2)
library(dplyr)
```

# 2) Import the data in R

We want you to get some practice importing data tables in R. As we saw in lecture, there
are multiple ways for reading in tables in R. One major approach consists of using R base
functions: read.table() and friends—e.g. read.csv(), read.delim(). Another major
approach, recently introduced, is the one provided by the family of functions in the package
"readr".

You will have to import the data using both approaches. This means that you are going to
import the data twice, but this is just for practicing purposes, and so that you can compare
both importing styles ("base" vs "readr").

What you should include in your Rmd is two code chunks with the commands to import
the data, one chunk with read.csv(), and the other with read_csv(). 

Create one data frame for the output of read.csv(), and use str() to display its structure. Likewise, create
another data frame for the output of read_csv() and use str() to check its structure.

In both cases you have to explicitly specify the data-type for each column as follows:

• the columns Player, Team, and Experience have to be declared as type character.
• the column Position has to be declared as a factor with levels 'C', 'PF', 'PG',
'SF', 'SG'.
• the column Salary has to be declared as type double (or real).
• the rest of the 19 columns have to be declared as type integer.

```{r}

colClasses <- c(
    'character',
    'character',
    'factor',
    'character',
    'real',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer',
    'integer'
)
dat1 <- read.csv(csv,colClasses = colClasses, stringsAsFactors = FALSE, na.strings = "N/A")
str(dat1)
dat1
```



```{r}
dat2 <- read_csv('nba2017-player-statistics.csv')
str(dat2)
```

```{r}
dat2 <- read_csv('nba2017-player-statistics.csv', col_types = list(
     Position = col_factor(c('C', 'PF', 'PG', 'SF', 'SG')),
     Salary = col_double()
))
str(dat2)
```

• recall that read.csv() uses the argument colClasses to specify data types.
• recall that read_csv() uses the argument col_types to specify data types.
In case you need, here’s a couple of resources for importing data (if you google about this
topic you’ll find more links):
• https://www.r-bloggers.com/using-colclasses-to-load-data-more-quickly-in-r/
• https://cran.r-project.org/web/packages/readr/vignettes/readr.html

# 3) Right after importing the data
Once you have the data in R, do a bit of preprocessing on the column Experience. This
column should be of type character because of the presence of the R values that indicate
rookie players.
Replace all the occurrences of "R" with 0, and then convert the entire column into integers.
```{r}
typeof(dat1$Experience)
dat1$Experience[dat1$Experience == "R"] <- "0"
dat1$Experience <- as.integer(dat1$Experience)
typeof(dat1$Experience)
```

# 4) Performance of players
As we mention above, in this assignment you will take into account other variables rather than
just the points scored by each player. More precisely, you are going to consider basic individual statistics commonly used in the NBA in order to get a proxy of a player’s performance.
Performance of NBA players can be measured in various ways. Perhaps the most popular
performance measure is known as the “Efficiency” statistic, simply referred to as EFF
https://en.wikipedia.org/wiki/Efficiency_(basketball)
EFF computes performance as an index that takes into account basic individual statistics:
points, rebounds, assists, steals, blocks, turnovers, and shot attempts (per game). It is derived
by a simple formula:
EFF = (PTS + REB + AST + STL + BLK - Missed FG - Missed FT - TO) / GP

EFF: efficiency
• PTS: total points
• REB: total rebounds
• AST: assists
• STL: steals
• BLK: blocks
• Missed FG: missed field goals
• Missed FT: missed free throws
• TO: turnovers
• GP: games played


In case you are curious, you can find more information about the player statistics and related
acronyms in the following wikipedia entry:
https://en.wikipedia.org/wiki/Basketball_statistics

To compute EFF, you will have to add the following variables to your data frame:
• Missed_FG (missed field goals) +
• Missed_FT (missed free throws) +
• PTS (total points)
• REB (total rebounds: offensive and defensive)
• MPG (minutes per game; NOT to be used when calculating EFF)


```{r}
dat1$Missed_FG <- dat1$FGA-dat1$FGM
dat1$Missed_FT <- dat1$FTA-dat1$FTM
dat1$PTS <-dat1$Points2*2+dat1$Points3*3+dat1$FTM*1
dat1$REB <- dat1$OREB+dat1$DREB
dat1$MPG <- dat1$MIN/dat1$GP
select(dat1, Player, Missed_FG, Missed_FT, PTS, REB, MPG, TO)
```


Once you have all the necessary statistics, add a variable EFF to the data frame using the
formula provided above.
EFF = (PTS + REB + AST + STL + BLK - Missed FG - Missed FT - TO) / GP

```{r}
dat1$EFF = (dat1$PTS + dat1$REB + dat1$AST + dat1$STL + dat1$BLK - dat1$Missed_FG - dat1$Missed_FT - dat1$TO) / dat1$GP
summary(dat1$EFF)
hist(dat1$EFF, col = 'grey', main = "EFF", xlab = "EFF", ylim = c(0, 200), xlim = c(-10, 40))
```
Compute summary statistics for EFF and confirm that you have the following results, as well
as a similar histogram:
Min. 1st Qu. Median Mean 3rd Qu. Max.
-0.600 5.452 9.090 10.140 13.250 33.840


• Display the player name, team, salary, and EFF value of the top-10 players by EFF in
decreasing order (display this information in a data frame).
```{r}
top <- select(slice(arrange(dat1, desc(EFF)), 1:10), Player, Team, Salary, EFF)
is.data.frame(top)
top
```

• Provide the names of the players that have a negative EFF.
```{r}
dat1$Player[which(dat1$EFF < 0)]
select(filter(dat1, dat1$EFF < 0), Player)
```

• Use the function cor() to compute the correlation coefficients between EFF and all the
variables used in the EFF formula.
```{r}
cor(dat1$EFF, dat1$PTS)
cor(dat1$EFF, dat1$REB)
cor(dat1$EFF, dat1$AST) 
cor(dat1$EFF, dat1$STL)
cor(dat1$EFF, dat1$BLK)
cor(dat1$EFF, dat1$Missed_FG)
cor(dat1$EFF, dat1$Missed_FT)
cor(dat1$EFF, dat1$TO)
```
• Notice that Missed_FG, Missed_FT, and TO contribute negatively to EFF, so make
sure to take into account this negative association when calculating the correlation
coefficients.
```{r}
cor(dat1$EFF, dat1$Missed_FG)*(-1)
cor(dat1$EFF, dat1$Missed_FT) *(-1)
cor(dat1$EFF, dat1$TO)*(-1)
```

• Display the computed correlations in descending order, either in a vector or a data
frame. And create a barchart with the correlations (bars in decreasing order) like the
one below.

```{r}
Name <- c("PTS", "REB", "AST", "STL", "BLK", "FG", "FT", "TO")
Corr <- c(
  cor(dat1$EFF, dat1$PTS),
 cor(dat1$EFF, dat1$REB),
cor(dat1$EFF, dat1$AST),
cor(dat1$EFF, dat1$STL),
cor(dat1$EFF, dat1$BLK),
cor(dat1$EFF, dat1$Missed_FG)*(-1),
cor(dat1$EFF, dat1$Missed_FT)*(-1),
cor(dat1$EFF, dat1$TO)*(-1)
)
CORR <- data_frame(Name,Corr)
CORR
answer <- arrange(CORR, desc(CORR$Corr))
answer
?barplot
barplot(answer$Corr, names.arg = answer$Name, 
        main = "Correlation between Player Stat and EFF",
        width = 2, ylim = c(-1, 1), 
        col = ifelse(answer$Corr > 0, "grey", "lightgreen"))
abline(h=0)
```



# 5) Efficiency and Salary

Once you’ve calculated the Efficiency statistic, produce a scatterplot between Efficiency
(x-axis) and Salary (y-axis), including a lowess smooth line (locally weighted scatterplot
smoothing). 
```{r}
ggplot(dat1, aes(x = EFF, y = Salary)) +
  geom_point() + 
  geom_smooth(method = loess)
```

Also, compute the linear correlation coefficient between them.
```{r}
cor(dat1$EFF, dat1$Salary)
```
What can you say about the relationship between these two variables?
*It is fairly positive. As EFF increases, salary increases as well.*

One aspect that has an important contribution in a player’s salary has to do with the years
of experience playing in the NBA. In fact, salaries of rookie players follow a special scale, as
explained by Tom Ziller in this post.
Why NBA players get paid so much more than NFL stars
Because rookie players (and other low-experience players) seem to form a different universe
of players, let’s see what’s happening with those individuals with a more “solid” or more
established trajectory in the NBA.
• Taking into account the column MPG (minutes per game) select those players that have
an MPG value of 20 or more minutes per game.
```{r}
players2 <- filter(dat1, MPG >=20)
select(players2, Player)
```
• Create a data frame players2 with these players.
```{r}
players2
```
• Use this data frame to create a scatterplot between Efficiency and Salary, including a
lowess smooth line.
```{r}
ggplot(players2, aes(x = EFF, y = Salary)) + 
  geom_point() + 
  geom_smooth(method = loess)
```
• Compute the linear correlation coefficient between these variables.
```{r}
cor(players2$EFF, players2$Salary)
```
• What can you say about the relationship between these two variables for the set of
“more established players”?
*It is lower for more experienced players than for less experiienced maybe because more important factors come into play or maybe salaries cannot go much higher.* 

In future assignments, we’ll consider other research questions, as well as other ways to explore
more variables, and perform a deeper multivariate analysis.


# 6) Comments and Reflections
Reflect on what was hard/easy, problems you solved, helpful tutorials you read, etc.

Question | Answer
-------------------------------------------------------------------------- | ----------------------------------------------
• What things were hard, even though you saw them in class/lab? | Still struggling with creating vectors.
• What was easy(-ish) even though we haven’t done it in class/lab? | Adding columns
• Did you need help to complete the assignment? If so, what kind of help? | No, just piazza
• How much time did it take to complete this HW? | More than 7 hours!
• What was the most time consuming part? | Figuring out all the basketball terms
• Was there anything that you did not understand? or fully grasped? | Need to see the answers and check correctness first.
• Was there anything frustrating in particular? | Basketball!
• Was there anything exciting? Something that you feel proud of? (Don’t be shy, we won’t tell anyone). | No. But I like ggplot.
















