---
title: "hw03-alina-skripets"
author: "Alina Skripets"
date: "October 10, 2017"
output: github_document
---

#About the Research Question

In this assignment you will focus on a hypothetical question: If you had to come up with
a ranking system for the teams, how would you rank them?

To make things more interesting, let’s pretend that the NBA does not work the way it does.
Let’s also pretend that the only available data is the player statistics, and nothing else. In
other words, we don’t know the number of wins (and losses) of each team, or which team
won the championship. Moreover, let’s assume there is no such championship. All we have is
the information about the players, and the goal is to find a ranking for the teams.

If these assumptions and the ranking idea seem awkward, think about the ranking systems
of universities, the ranking of companies in a certain industry, or the ranking of countries
according to some economic or socio-demographic indicators (see examples below):
• U.S. News National University Rankings
• U.S. News Overall Best Countries Ranking
• Fortune Tech The 30 Best Workplaces in Technology

In this assignment, you are going to consider different ways to rank the NBA teams. From
simple rankings based on a given observed variable, to rankings based on derived indices
like efficiency (i.e. EFF), to rankings based on a composite index using Principal Component
Analysis (PCA).

#Ranking of Teams

The analysis stage of this assignment has to do with looking at various ways to rank the
teams. Use an Rmd file for this part of your project.

#Basic Rankings
Start by ranking the teams according to salary, arranged in decreasing order. Use ggplot()
to create a barchart (horizontally oriented), like the one shown below. The vertical red line
is the average team salary. 
You will have to look at the following resources to learn how to obtain such type of ggplot.
• Horizontal barplot in ggplot
https://stackoverflow.com/questions/10941225/horizontal-barplot-in-ggplot2
• axis labels in ggplot2
http://ggplot2.tidyverse.org/reference/labs.html

```{r}
rm(list=ls())
library(ggplot2)
library(readr)
library(dplyr)
data <- read_csv('hw03/data/nba2017-teams.csv')
#as.factor(data$team)
```


```{r}
h1=mean(data$salary)
ggplot(data, aes(x = reorder(team, salary), y = salary)) + 
  geom_bar(stat='identity', fill = 'grey') + 
  coord_flip() + 
  geom_abline(aes(slope = 0, intercept = h1), color = 'red', size = 2, alpha = 0.5) + 
  ggtitle("NBA Teams ranked by Total Salary") +
  labs(y = "Salary (in millions)", x = "Team")
```

Create another bar chart of teams ranked by total points. The vertical red line is the average team points.

```{r}
h2=mean(data$points)
ggplot(data, aes(x = reorder(team, points), y = points)) + 
  geom_bar(stat='identity', fill = 'grey') + 
  coord_flip() + 
  geom_abline(aes(slope = 0, intercept = h2), color = 'red', size = 2, alpha = 0.5) + 
  ggtitle("NBA Teams ranked by Total Points") +
  labs(y = "Total Points", x = "Team")
```

Use efficiency to obtain a third kind of ranking, and create an associated bar chart of
teams ranked by total efficiency. The vertical red line is the average team efficiency.

```{r}
h3=mean(data$efficiency)
ggplot(data, aes(x = reorder(team, efficiency), y = efficiency)) + 
  geom_bar(stat='identity', fill = 'grey') + 
  coord_flip() + 
  geom_abline(aes(slope = 0, intercept = h3), color = 'red', size = 2, alpha = 0.5) + 
  ggtitle("NBA Teams ranked by Total Efficiency") +
  labs(y = "Total Efficiency", x = "Team")
```


#Provide concise descriptions of the obtained rankings so far.

Ranking provide similar information. Shows that all the criteria combined together matter.

#Principal Components Analysis (PCA)
Perform a principal components analysis (PCA) on the following variables, to use the first
principal component (PC1) as another index to rank the teams:
• points3
• points2
• free_throws
• off_rebounds
• def_rebounds
• assists
• steals
• blocks
• turnovers
• fouls
Use prcomp()—NOT to confuse with princomp()—to perform a PCA, specifying the argument
scale. = TRUE (i.e. PCA on standardized data).

```{r}
a <- cbind(data$points3, data$points2, data$free_throws, data$off_rebounds, data$def_rebounds, data$assists, data$steals, data$blocks, data$turnovers, data$fouls)
a <- as.data.frame(a)
pc <- prcomp(a, scale.=TRUE, center = TRUE)
pc <- pc[[5]][ ,1:2]
fin <- cbind(data$team, pc)
fin  <- as.data.frame(fin)
colnames(fin) <- c('Team', 'PC1', 'PC2')
fin$PC1 <- as.numeric(as.character(fin$PC1))
fin$PC2 <- as.numeric(as.character(fin$PC2))
str(fin)
ggplot(fin, aes(x = PC1, y=PC2)) + 
  geom_text(aes(label = Team)) + 
  ggtitle("PCA Plot (PC1 and PC2)") + 
geom_abline(intercept=0, slope=0) +
  geom_abline(intercept = 0, slope=999999999)
```

## Index based on PC1

In order to build an index based on the first PC, you are going to transform PC1. To get a
more meaningful scale, you can rescale the first PC with a new scale ranging from 0 to 100.
Let z1 be the first principal component. The transformed score s1, ranging on a scale from 0
to 100, can be obtained as:
s1 = 100 ×
z1 − min(z1)
max(z1) − min(z1)
```{r}
s1 <- 100*((fin$PC1-min(fin$PC1))/(max(fin$PC1)-min(fin$PC1)))
s1
finfin <- data.frame(s1=s1, team=data$team)
finfin
```
Once you have obtained the rescaled PC1, you can produce a barchart like the previous ones:

```{r}
ggplot(finfin, aes(x = reorder(team, s1), y = s1)) + 
  geom_bar(stat='identity', fill = 'grey') + 
  coord_flip() + 
  geom_abline(aes(slope = 0, intercept = h3), color = 'red', size = 2, alpha = 0.5) + 
  ggtitle("NBA Teams ranked by Total Efficiency") +
  labs(y = "First PC (scaled from 0 to 100)", x = "Team")
```

We can see that GSW came out on the top according to the new rating. 
This rating should be more reliable to judge how good the teams are because it combines all the criteria in one. The results are still similar to those we obtained in our previous analysis. 


##Comments and Reflections 
Reflect on what was hard/easy, problems you solved, helpful tutorials you read, etc.   
Was this your first time working on a project with such file structure? If yes, how do
you feel about it?   
Yes, it wasn't hard until it came up on the midterm.  
Was this your first time using relative paths? If yes, can you tell why they are important
for reproducibility purposes?   
Yes, and yes, they are much easier.    
Was this your first time using an R script? If yes, what do you think about just writing
code?   
No, I used it in the lab before.    
What things were hard, even though you saw them in class/lab?   
It was okay, just took some creative thinking and lots of time.    
What was easy(-ish) even though we haven’t done it in class/lab?   
Everything was okay, not that hard.  
Did anyone help you completing the assignment? If so, who?   
No  
How much time did it take to complete this HW?  
LONG TIME    
What was the most time consuming part?  
PCA naturally and understanding it.   
Was there anything interesting?   
The purpose of PCA was really interesting.   





















