---
title: "WarmUp2"
author: "Alina Scripts"
date: "September 11, 2017"
output: github_document
---

Lets download a file to my computer's working directory first and upload it to R working space.

```{r}
rm(list=ls())
github <- 'https://github.com/ucb-stat133/stat133-hws-fall17/'
repo <- 'raw/master/warmup02/data/nba2017-salary-points.RData'

download.file(url = paste0(github, repo), 
destfile = "nba2017-salary-points.RData")

```

```{r}
rm(list=ls())
load("nba2017-salary-points.RData")
```

Now we list all the variables that we have in the data package. We notice that because I am not very good at R, I have two extra objects: github and repo.Or not, I fixed it using rm function.

```{r}
# list the available objects
ls()
```

Now we hapazardly try to find out that type of salary is ' `r typeof (salary) ` ', whether or not team is a vector - `r is.vector(team) `, whether or not position object is a list -  `r is.list(position) ` and what is the class of player object, which is' `r class(player) ` '    

We successfully used indented code! But I still do't understand what the professor meant by finding out *flavor*.

Now we check that all the objects are the same length:

```{r}

length(salary)
length(team)
length(position)
length(player)
length(points)
length(experience)
length(points1)
length(points2)
length(points3)
```
They are the same length!
On to the next part.


# Quantitative variable: *Salary* 

```{r}
summary(salary)
```
```{r}
mean(salary)
sd(salary)
min(salary)
max(salary)
median(salary)
quantile(salary, probs = seq (0, 1, 0.25))
```

I can find all the statistics above, no problem. But how do i find typical value/ mode? I have no idea.

```{r}
?tabulate
getmode <- function(salary) {
  uniqv <- unique(salary)
  uniqv[which.max(tabulate(match(salary,uniqv)))]
}
getmode(salary)
```
Wait, I did it!
Now I can also find a spread by substracting min from max.

```{r}
spread= max(salary)-min(salary)
spread
```
Oooo, what do we have here. 

```{r}
hist(salary)
boxplot(salary)
```
/

/**What is the overall shape?** This histogram is scewed to the right with th bulk of salaries in the first bin. 

**What is the typical center?** Typical center is at the left.

**Overall range, along with an interval of typical measurements.** Graphs have a huge range and typical interval is small because we have lots of data.

**Try to explain why your observations are important or interesting.** Observation is interesting because it proves that while the mean of the distribution is pretty high, it doesn't mean that players earn a lot on average. Median here is a better description of data.

#How do I creat a density polygon? No idea :(


```{r}
??density
y <- c(density(salary))

#cdplot(salary)
```

# Catigorical variable: *Team*

Is team variable a factor? `r is.factor(team) `
Lets also find out how many variable are in each category aka frequency of each team.

```{r}
teamfreq <- table(team)
teamfreq
```

Based on this, we can calculate proportions or relative frequencies of each team. 

```{r}
prop.table(teamfreq)
```

```{r}
barplot(prop.table(teamfreq))
```
/

**What is the overall shape?** Pretty much flat.

**What is the typical center?** Each team is approximately the same size.

**Overall range, along with an interval of typical measurements.** 14 or 15 players in eah team.

**Try to explain why your observations are important or interesting.** Clearly, game rules and team size regulations dictated the shape of the barplot.



# Reflections

Question|Answer
-------------------------------------------------------------|-------------------------------------------------------------
What things were hard, even though you saw them in class? | Finding a mode.
What was easy(-ish) even though we haven't done it in class? | It was easy to figure out indented code.
What type of "errors" you struggled with (if any)? | I still can't figure out density polygon.
What are the parts you are not fully understanding? | Falvor of an object and density polygon creation. Creating my own function for mode was hard too.
What was the most time consuming part? | Comments.
Did you collaborate with other students? If so, with who? In what manner? | No, Piazza was unresponsive. 
Was there any frustrating issue? (e.g. RStudio cryptic error, one or more package not playing nice) | Could get the ls() function to not list github and repo as separate elements.



##For people who want to take things further. 

How to use the R package "xtable" to produce nicely formatted tables.
```{r}
#install.packages("xtable")

```
```{r}
library("xtable")
```




```{r}
?par
plot(points,salary,type="l", col= "light blue",lwd = 5)
```


